Approximation as Differentiation

### 2025-07-08 main/approx-as-tangents.tex
abramsky-jung.
Missing \mathrm.
Citations.

We motivate our approach to \GPS by showing how to combine ideas from differential geometry and stable domain theory to reconstruct \GPS in a denotational setting. From this basis, we apply the CHAD framework of V{\'a}k{\'a}r \etal to obtain a denotational model of automatic differentiation that gives us an effective way to interpret programs as functions along with their forward and backwards approximation maps. \bob{and we did it in Agda, so we can run it}

@@

Domain theory is a method for defining the semantics of programs that
handle infinite data such as functions or infinite streams. Domains
are certain partially ordered sets where the ordering denotes a
relationship of qualitative information content: if $x \sqsubseteq y$,
then $y$ may contain more information than $x$. For example, if $x$
and $y$ are functions, then $y$ may be defined at more values than
$x$. Infinite objects are understood in terms of their approximations
in this sense, and domains are assumed to be closed under least upper
bounds (lubs) of directed sets, meaning that any internally consistent
collection of elements has a ``completion'' that contains all the
information covered by the set. Programs are interpreted as monotone
functions that preserve directed lubs. Monotonicity captures the idea
that if the input gets more defined, then the output can get more
defined. Preservation of lubs, or \emph{continuity}, states that a
function interpreting a program cannot act inconsistently on
approximations and their completion, which corresponds to the
intuitive idea that a function that is computable cannot look at a
non-finite amount of input to produce a finite output. \cite{abramsky-jung} is a comprehensive introduction to domain theory.

@@

\begin{proof}
  \AGDA. See \citet{amadio-curien} (FIXME: theorem reference) for the
  case when the posets are also cpos. The crucial technical step,
  identified by \citet{berry79}, is that the ordering on conditionally
  multiplicative functions is not the extensional ordering
  ($f \sqsubseteq_{\mathrm{ext}} g$ iff
  $\forall x. f(x) \sqsubseteq g(x)$) but instead the stable ordering:
  $f \sqsubseteq_{\mathrm{st}} g$ iff $f \sqsubseteq_{ext} g$ and
  $\forall x, x'.\,x \sqsubseteq x' \Rightarrow f(x) = f(x') \land
  g(x)$.
\end{proof}

@@

  Relatedly, \citet{edalat-heckmann98} proposed a Scott domain of
  formal balls on a metric space, where again the maximal elements are
  the points of the original space. More generally,
  \citet{continuous-lattice-book} (Section V-6) describes \emph{Domain
    environments}, which are domains whose maximal elements are
  exactly the points of a topological space. We are not aware of any
  work linking domain environments in general to stable functions. It
  would be interesting to see whether the situation for intervals,
  where approximations must be relative to a nominated point, repeats
  in general when considering conditionally multiplicative functions.
\end{example}

>>>>

We motivate our approach to \GPS by showing how to combine ideas from differential geometry and stable domain theory to reconstruct \GPS in a denotational setting. From this basis, we apply the CHAD framework of V{\'a}k{\'a}r \etal to obtain a denotational model of automatic differentiation that gives us an effective way to interpret programs as functions along with their forward and backwards approximation maps.

@@

Domain theory is a method for defining the semantics of programs that
handle infinite data such as functions or infinite streams. Domains
are certain partially ordered sets where the ordering denotes a
relationship of qualitative information content: if $x \sqsubseteq y$,
then $y$ may contain more information than $x$. For example, if $x$
and $y$ are functions, then $y$ may be defined at more values than
$x$. Infinite objects are understood in terms of their approximations
in this sense, and domains are assumed to be closed under least upper
bounds (lubs) of directed sets, meaning that any internally consistent
collection of elements has a ``completion'' that contains all the
information covered by the set. Programs are interpreted as monotone
functions that preserve directed lubs. Monotonicity captures the idea
that if the input gets more defined, then the output can get more
defined. Preservation of lubs, or \emph{continuity}, states that a
function interpreting a program cannot act inconsistently on
approximations and their completion, which corresponds to the
intuitive idea that a function that is computable cannot look at a
non-finite amount of input to produce a finite output. \citet{abramsky-jung} provide a comprehensive introduction to domain theory.

@@

\begin{proof}
  \AGDA. See \citet{amadio-curien} (FIXME: theorem reference) for the
  case when the posets are also cpos. The crucial technical step,
  identified by \citet{berry79}, is that the ordering on conditionally
  multiplicative functions is not the extensional ordering
  ($f \sqsubseteq_{\mathrm{ext}} g$ iff
  $\forall x. f(x) \sqsubseteq g(x)$) but instead the stable ordering:
  $f \sqsubseteq_{\mathrm{st}} g$ iff $f \sqsubseteq_{\mathrm{ext}} g$ and
  $\forall x, x'.\,x \sqsubseteq x' \Rightarrow f(x) = f(x') \land
  g(x)$.
\end{proof}

@@

  Relatedly, \citet{edalat-heckmann98} proposed a Scott domain of
  formal balls on a metric space, where again the maximal elements are
  the points of the original space. More generally,
  \citet{continuous-lattice-book} (Section V-6) describe \emph{domain
    environments}, which are domains whose maximal elements are
  exactly the points of a topological space. We are not aware of any
  work linking domain environments in general to stable functions. It
  would be interesting to see whether the situation for intervals,
  where approximations must be relative to a nominated point, repeats
  in general when considering conditionally multiplicative functions.
\end{example}
